Repeatability of Pre-Flashover Fire Patterns on Gypsum Wall Board

### 2015-03-17 Projects/Madrzykowski_Thesis/Report/Madrzykowski_Thesis.tex
Madrzy_Thesis: minor latex formatting

The ability to recreate or simulate a fire pattern or certain characteristics of a fire pattern was examined with a variety of tools.  The first characteristic examined was the ability to determine based on the calculated heat flux to the wall if a "visible or measurable physical change" would occur to the painted paper surface of the gypsum wallboard.  The next characteristic of the fire pattern that was examined was the height.  The height of the fire pattern was then compared to the predictive results from flame height correlations. Heskestad developed an empirical method for calculating the mean visible flame height for fuels which do not have substantial in-depth combustion~\cite{Beyler:1986,Heskestad:SFPE}. This correlation was compared with the flame height results of the source fires and the instrumented wall experiments.

@@

The natural gas burner was 30.5 cm on a side and the top surface of the burner was 7.5~cm above the floor. The shell of the burner was made from steel and filled with ``pea sized'' gravel with an average diameter of 6~mm.  A steady state heat release rate of approximately 80 kW was used for the experiments.

@@

A 35~mm single lens reflex camera was used to record digital still photographs of the fires.  The camera had an 8 megapixel resolution.  A mini DV camcorder with 0.7 megapixel resolution was used to capture the video.  A firefighting type thermal imager with 0.08 megapixel resolution was used to capture infrared images in the 8 to 14 micron range.  Each camera was positioned on a tripod with a distance of 4.5 m between the centerline of burner to camera lens 4.5~m.  The height of each camera was adjusted so that the center of camera lens was positioned 0.5 m above top of the burner.

@@

On each fuel’s averaged heat release rate curve, a bar representing the 95~\% confidence level ($2\sigma$), based on Type A evaluation of the uncertainty is provided~\cite{Taylor:1994}.  A Type A evaluation of the uncertainty is any statistical analysis of the data from a series of observations.  In this case the evaluation assumed a normal distribution of the data such that doubling the calculated standard deviation (2σ) would yield a confidence level of approximately 95~\%.

@@

For the natural gas experiments, approximately 30~s of ``steady state'' video were captured.  In the cases of the gasoline and polyurethane fueled experiments, the video capture period bounded the peak heat release rate.  In the case of the gasoline fueled fires, the peaks occurred over a long period of time, so a 30 s video capture could be made which was representative of the peak burning rate.  The peak duration for the polyurethane foam fires was shorter so, only a 15~s video capture was used.

@@

A free standing steel frame wall with a 12.7 mm thick, 2.44 m high by 1.22 m wide, calcium silicate face was instrumented to measure the thermal exposure of the wall from the fire in terms of temperature and heat flux.  Two vertical arrays of water cooled, Schmidt-Boelter total heat flux gauges, as previously described in the instrumentation section above, were installed at 20 cm intervals from 0.2 m to 1.2 m above the burner.  One vertical array was centered on the burner and the other vertical array was centered on the right edge of the burner. The arrangement is shown in Fig.~\ref{Instrumented_NC_Wall}. Bare-bead, Chromel-Alumel (type K) thermocouples, with a 0.5~mm nominal diameter were installed next to the center of each heat flux sensors and continued, at 20 cm intervals, up to a height of 2.2 m above the surface of the burner. Photographs of the front and rear of the instrumented wall are shown in Figures~\ref{Instrumented_Wall_Front_photo} and \ref{Instrumented_Wall_Rear_photo} The thermocouples where installed in small holes from the back of the wall of the wall.  The bead of the thermocouple was flush with the surface on the calcium silcate board and was in contact with the board, as shown in Figure~\ref{Instrumented_Wall_Close_up_TC_HF_photo} For the center array,a second thermocouple was installed on the backside of the calcium silicate board, approximately 10 mm below the TC positioned on the front surface of the wall. The photograph in Figure~\ref{Instrumented_Wall_Detail_Rear_TC_photo} shows a rear thermocouple with the bead in contact with the rear surface of the wall and the relative position to the thermocouple on the front surface of the wall. In a manner similar to the source fire characterization experiments, the instrumented wall experiments were conducted under the NIST Large Fire Research Laboratory’s 3~m by 3~m oxygen depletion calorimeter.

@@

The cone calorimeter is one of the most commonly used bench-scale rate of heat release apparatus that uses the oxygen consumption method. It was developed at NIST in the 1980s~\cite{babrauskas:1984}. The cone calorimeter has been adopted as ASTM E1354, Test Method for Heat and Visible Smoke Release Rates for Materials and Products Using an Oxygen Consumption Calorimeter~\cite{ASTM E1354}. The cone calorimeter consists of a heater, spark ignitor, sample holder, and load cell located underneath an exhaust hood. Typically, the sample is located in the open with free access of air to the combustion zone. The heater consists of a 5~kW electrical heating element inside an insulated stainless-steel conical shell. Samples can be tested in either a horizontal or vertical orientation. When tests are performed in the horizontal configuration, the specimen is positioned approximately 25~mm beneath the bottom plate of the cone heater. Flames and products of combustion pass through a circular opening at the top of the heater. The heater can expose samples to a maximum irradiance of approximately 100~kW/m\$\^2\$. Figure  is a schematic of the cone calorimeter.

For piloted ignition tests, an electric spark ignitor is positioned at the top of vertical samples and over the center of horizontal samples. Samples are typically 100~mm squares. Combustion products and dilution air are extracted through the hood and exhaust duct by a high temperature fan. The flow rate can be adjusted between 0.01 and 0.03 m\$\^3\$/s. The volumetric flow rate is kept constant during testing. The sample is mounted on a load cell to determine mass loss rate during a test, and smoke obscuration is measured using a laser light source. The gas flow rate in the exhaust duct is calculated from the pressure drop across and temperature at an orifice plate in the duct. Finally, the concentrations of oxygen, carbon dioxide, carbon monoxide, were measured using appropriate instruments. Heat release rate is calculated from the gas concentration and mass flow measurements.

@@

The radiant flooring panel test method is known as ASTM~E648, Standard Method of Test for Critical Radiant Flux of Floor-Covering Systems Using a Radiant Heat Energy Source~\cite{ASTM_E648}. A 1~m long test specimen is mounted horizontally beneath a pre-mixed air-gas-fueled radiant panel. The panel is inclined at 30~\$\^°\$, providing a heat flux on the test specimen varying from 10~kW/m\$\^2\$ at the point closest to the panel to 1~kW/m\$\^2\$ at the farthest point. A pilot flame is used to ignite the test sample. The objective of the test is to determine the critical radiant flux for a given sample.  The critical radiant flux corresponds to the flux at the point of maximum flame propagation, or the point on the sample where the flame spread stopped. For this study the samples examined were cut from painted gypsum wallboard panels that were prepared with the panels used in the compartment experiments.

@@

The fire pattern experiments were conducted in residential scale compartments built for the purpose of these experiments.  The compartments were constructed on top of a portable wood floor platform.  Each of the test compartments consisted of three 3.6~m long by 2.4~m high wood framed walls. The wood elements of the frame were composed of kiln-dried fir with a nominal cross section of 90 mm by 38 mm.  A partial ceiling with a width of at least 1.2 m, measured from the back wall of the compartment was installed across the width of the compartment, this allowed for the plume to impact the ceiling and form a ceiling jet along the back wall, before the fire gases flowed up and out of the compartment and into the 6~m by 6~m hood in the NIST Large Fire Research Laboratory. The interior surface of the walls and ceiling was composed of 12.7 mm thick, regular core gypsum wallboard. As defined by ASTM, gypsum wallboard is designed for use on walls, ceilings, or partitions.  It is composed of a non-combustible core, essentially gypsum, with paper bonded to the surface of the core. The back wall of the compartment consisted of three full gypsum board panels, each 1.2 m wide and 2.4 m high.  The front wall of the compartment was open along the full width of the compartment to minimize the impact of entrained air being channeled through a doorway. These experiments were conducted without instrumentation in the walls for several reasons; to compare the non-instrumented gypsum wallboard fire pattern with the instrumented gypsum wallboard fire pattern results to determine if the instrumentation affected the pattern generation, to provide more realistic conditions, and to enable the conduct of additional replicate experiments given the limited laboratory availability.  Based on the outcome of the instrumented calibration target wall results, burner positions were chosen. After each burn, the fire patterns were measured, photographed and analyzed for repeatability. The fire damaged wallboard was subsequently removed and replaced with new sections of painted gypsum wallboard. The number of experiments for each case was dependent on the repeatability and similarity of the pattern development. Several types of fire pattern experiments were conducted in the compartments, as detailed in the sections to follow, to examine the impact of the insulation behind the gypsum wallboard panel on the resulting fire pattern and other series of experiments were conducted to compare the fire patterns based on the burner centered on the rear wall or in or near a corner on the rear wall of the compartment.

@@

 Impact of Wall Construction 
 Additional experiments were conducted with the natural gas and gasoline fires to examine the impact of the method of construction of the walls on the development of the fire patterns.  Three cases were examined: 1) Open back, 12.7 mm thick, regular core, painted gypsum board on the front of a wood frame only, 2) Closed back, 12.7mm thick, regular core, painted gypsum board on the front and back of a wood frame representative of an interior wall, and 3)	Insulated,12.7 mm thick, regular core, painted gypsum board on the front and back of a wood frame, with fiberglas batt insulation in the voids of the wood frame representative of an exterior wall.

The fuel or burner was centered on the middle panel of the rear wall.  As in the fire source characterization experiments the top of the fuel in each case was positioned 7.5 cm above the floor of the compartment.  Each of the painted gypsum board panels that were used for a fire pattern had a 5 cm grid drawn on it to assist with the determination of the geometry of the fire pattern.

>>>>

The ability to recreate or simulate a fire pattern or certain characteristics of a fire pattern was examined with a variety of tools.  The first characteristic examined was the ability to determine based on the calculated heat flux to the wall if a ``visible or measurable physical change'' would occur to the painted paper surface of the gypsum wallboard.  The next characteristic of the fire pattern that was examined was the height.  The height of the fire pattern was then compared to the predictive results from flame height correlations. Heskestad developed an empirical method for calculating the mean visible flame height for fuels which do not have substantial in-depth combustion~\cite{Beyler:1986,Heskestad:SFPE}. This correlation was compared with the flame height results of the source fires and the instrumented wall experiments.

@@

The natural gas burner was 30.5~cm on a side and the top surface of the burner was 7.5~cm above the floor. The shell of the burner was made from steel and filled with ``pea sized'' gravel with an average diameter of 6~mm.  A steady state heat release rate of approximately 80 kW was used for the experiments.

@@

A 35~mm single lens reflex camera was used to record digital still photographs of the fires.  The camera had an 8 megapixel resolution.  A mini DV camcorder with 0.7 megapixel resolution was used to capture the video.  A firefighting type thermal imager with 0.08 megapixel resolution was used to capture infrared images in the 8 to 14 micron range.  Each camera was positioned on a tripod with a distance of 4.5~m between the centerline of burner to camera lens 4.5~m.  The height of each camera was adjusted so that the center of camera lens was positioned 0.5~m above top of the burner.

@@

On each fuel’s averaged heat release rate curve, a bar representing the 95~\% confidence level ($2\sigma$), based on Type A evaluation of the uncertainty is provided~\cite{Taylor:1994}.  A Type A evaluation of the uncertainty is any statistical analysis of the data from a series of observations.  In this case the evaluation assumed a normal distribution of the data such that doubling the calculated standard deviation (2$\sigma$) would yield a confidence level of approximately 95~\%.

@@

For the natural gas experiments, approximately 30~s of ``steady state'' video were captured.  In the cases of the gasoline and polyurethane fueled experiments, the video capture period bounded the peak heat release rate.  In the case of the gasoline fueled fires, the peaks occurred over a long period of time, so a 30~s video capture could be made which was representative of the peak burning rate.  The peak duration for the polyurethane foam fires was shorter so, only a 15~s video capture was used.

@@

A free standing steel frame wall with a 12.7~mm thick, 2.44~m high by 1.22~m wide, calcium silicate face was instrumented to measure the thermal exposure of the wall from the fire in terms of temperature and heat flux.  Two vertical arrays of water cooled, Schmidt-Boelter total heat flux gauges, as previously described in the instrumentation section above, were installed at 20 cm intervals from 0.2~m to 1.2~m above the burner.  One vertical array was centered on the burner and the other vertical array was centered on the right edge of the burner. The arrangement is shown in Fig.~\ref{Instrumented_NC_Wall}. Bare-bead, Chromel-Alumel (type K) thermocouples, with a 0.5~mm nominal diameter were installed next to the center of each heat flux sensors and continued, at 20~cm intervals, up to a height of 2.2~m above the surface of the burner. Photographs of the front and rear of the instrumented wall are shown in Figures~\ref{Instrumented_Wall_Front_photo} and \ref{Instrumented_Wall_Rear_photo} The thermocouples where installed in small holes from the back of the wall of the wall.  The bead of the thermocouple was flush with the surface on the calcium silcate board and was in contact with the board, as shown in Figure~\ref{Instrumented_Wall_Close_up_TC_HF_photo} For the center array,a second thermocouple was installed on the backside of the calcium silicate board, approximately 10~mm below the TC positioned on the front surface of the wall. The photograph in Figure~\ref{Instrumented_Wall_Detail_Rear_TC_photo} shows a rear thermocouple with the bead in contact with the rear surface of the wall and the relative position to the thermocouple on the front surface of the wall. In a manner similar to the source fire characterization experiments, the instrumented wall experiments were conducted under the NIST Large Fire Research Laboratory’s 3~m by 3~m oxygen depletion calorimeter.

@@

The cone calorimeter is one of the most commonly used bench-scale rate of heat release apparatus that uses the oxygen consumption method. It was developed at NIST in the 1980s~\cite{babrauskas:1984}. The cone calorimeter has been adopted as ASTM E1354, Test Method for Heat and Visible Smoke Release Rates for Materials and Products Using an Oxygen Consumption Calorimeter~\cite{ASTM E1354}. The cone calorimeter consists of a heater, spark ignitor, sample holder, and load cell located underneath an exhaust hood. Typically, the sample is located in the open with free access of air to the combustion zone. The heater consists of a 5~kW electrical heating element inside an insulated stainless-steel conical shell. Samples can be tested in either a horizontal or vertical orientation. When tests are performed in the horizontal configuration, the specimen is positioned approximately 25~mm beneath the bottom plate of the cone heater. Flames and products of combustion pass through a circular opening at the top of the heater. The heater can expose samples to a maximum irradiance of approximately 100~kW/m$^2$. Figure  is a schematic of the cone calorimeter.

For piloted ignition tests, an electric spark ignitor is positioned at the top of vertical samples and over the center of horizontal samples. Samples are typically 100~mm squares. Combustion products and dilution air are extracted through the hood and exhaust duct by a high temperature fan. The flow rate can be adjusted between 0.01 and 0.03~m$^3$/s. The volumetric flow rate is kept constant during testing. The sample is mounted on a load cell to determine mass loss rate during a test, and smoke obscuration is measured using a laser light source. The gas flow rate in the exhaust duct is calculated from the pressure drop across and temperature at an orifice plate in the duct. Finally, the concentrations of oxygen, carbon dioxide, carbon monoxide, were measured using appropriate instruments. Heat release rate is calculated from the gas concentration and mass flow measurements.

@@

The radiant flooring panel test method is known as ASTM~E648, Standard Method of Test for Critical Radiant Flux of Floor-Covering Systems Using a Radiant Heat Energy Source~\cite{ASTM_E648}. A 1~m long test specimen is mounted horizontally beneath a pre-mixed air-gas-fueled radiant panel. The panel is inclined at 30~$^{\circ}$, providing a heat flux on the test specimen varying from 10~kW/m$^2$ at the point closest to the panel to 1~kW/m$^2$ at the farthest point. A pilot flame is used to ignite the test sample. The objective of the test is to determine the critical radiant flux for a given sample.  The critical radiant flux corresponds to the flux at the point of maximum flame propagation, or the point on the sample where the flame spread stopped. For this study the samples examined were cut from painted gypsum wallboard panels that were prepared with the panels used in the compartment experiments.

@@

The fire pattern experiments were conducted in residential scale compartments built for the purpose of these experiments.  The compartments were constructed on top of a portable wood floor platform.  Each of the test compartments consisted of three 3.6~m long by 2.4~m high wood framed walls. The wood elements of the frame were composed of kiln-dried fir with a nominal cross section of 90~mm by 38~mm.  A partial ceiling with a width of at least 1.2~m, measured from the back wall of the compartment was installed across the width of the compartment, this allowed for the plume to impact the ceiling and form a ceiling jet along the back wall, before the fire gases flowed up and out of the compartment and into the 6~m by 6~m hood in the NIST Large Fire Research Laboratory. The interior surface of the walls and ceiling was composed of 12.7~mm thick, regular core gypsum wallboard. As defined by ASTM, gypsum wallboard is designed for use on walls, ceilings, or partitions.  It is composed of a non-combustible core, essentially gypsum, with paper bonded to the surface of the core. The back wall of the compartment consisted of three full gypsum board panels, each 1.2~m wide and 2.4~m high.  The front wall of the compartment was open along the full width of the compartment to minimize the impact of entrained air being channeled through a doorway. These experiments were conducted without instrumentation in the walls for several reasons; to compare the non-instrumented gypsum wallboard fire pattern with the instrumented gypsum wallboard fire pattern results to determine if the instrumentation affected the pattern generation, to provide more realistic conditions, and to enable the conduct of additional replicate experiments given the limited laboratory availability.  Based on the outcome of the instrumented calibration target wall results, burner positions were chosen. After each burn, the fire patterns were measured, photographed and analyzed for repeatability. The fire damaged wallboard was subsequently removed and replaced with new sections of painted gypsum wallboard. The number of experiments for each case was dependent on the repeatability and similarity of the pattern development. Several types of fire pattern experiments were conducted in the compartments, as detailed in the sections to follow, to examine the impact of the insulation behind the gypsum wallboard panel on the resulting fire pattern and other series of experiments were conducted to compare the fire patterns based on the burner centered on the rear wall or in or near a corner on the rear wall of the compartment.

@@

 Impact of Wall Construction
 Additional experiments were conducted with the natural gas and gasoline fires to examine the impact of the method of construction of the walls on the development of the fire patterns.  Three cases were examined: 1) Open back, 12.7~mm thick, regular core, painted gypsum board on the front of a wood frame only, 2) Closed back, 12.7~mm thick, regular core, painted gypsum board on the front and back of a wood frame representative of an interior wall, and 3)	Insulated, 12.7~mm thick, regular core, painted gypsum board on the front and back of a wood frame, with fiberglas batt insulation in the voids of the wood frame representative of an exterior wall.

The fuel or burner was centered on the middle panel of the rear wall.  As in the fire source characterization experiments the top of the fuel in each case was positioned 7.5~cm above the floor of the compartment.  Each of the painted gypsum board panels that were used for a fire pattern had a 5 cm grid drawn on it to assist with the determination of the geometry of the fire pattern.
